Genetic alterations and expression patterns of genes coding for FGF, FGFR, and FGFBP proteins in urothelial cancers

### 2024-12-07 report/markdown/report.Rmd
HPA data in the report and figures

* Five collectives of urothelial cancer patients were evaluated: the GENIE BLCA cohort, an urothelial carcinoma subset of the large GENIE project (n = `r nrow(genie$clinic)`) [@Sweeney2017], the predominantly muscle-invasive MSK IMPACT cohort (`r nrow(msk$clinic)`) [@Clinton2022], the muscle-invasive TCGA BLCA collective (n = `r nrow(tcga$clinic)`) [@Liu2018; @Robertson2017], the IMvigor cohort of late stage, platinum-refractory cancers subjected to an anti-PD-L1 treatment (`r nrow(imvigor$clinic)`) [@Balar2017], and the BCAN cohort of metastatic urothelial cancers [@Damrauer2022]. Detailed information on somatic mutations and their effects on protein sequence was provided for the GENIE BLCA, MSK IMPACT, TCGA BLCA, and BLCA cohorts. Binarized indexes of presence/absence of somatic mutations in selected genes were available for the IMvigor cohort. Binarized data on gene amplifications and deletions were obtained for the GENIE BLCA, MSK IMPACT, and TCGA BLCA collectives. Gene expression quantification by RNA sequencing and assignment to consensus molecular classes [@Kamoun2020] were retrieved for the TCGA BLCA, IMvigor, and BCAN cohorts. Overall survival information was available for the MSK IMPACT, TCGA BLCA, IMvigor, and BCAN cohorts; data on disease-specific and relapse-free survival was provided for the TCGA BLCA collective. Approximate overall survival was calculated for the GENIE BLCA project participants based on age at enrollment in years and follow-up time in days. Characteristic of the investigated collectives is presented in __Table \@ref(tab:tab-cohorts)__.

@@

@@

flextable::flextable(rep_tabs$cohorts) %>% 
    set_widths(c(2.5, rep(2.7, 5), 2.3, 2.2)) %>% 
  footnote(1, 7:8, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

## Co-regulation of genes coding for FGFR, FGF, and FGFBP proteins

@@

Five collectives of urothelial carcinoma patients were evaluated in the current report: 
the urothelial cancer subset of the GENIE cohort referred to as GENIE BLCA [@Sweeney2017], 
the MSK IMPACT collective [@Clinton2022], 
the bladder cancer TCGA collective of muscle-invasive urothelial cancers referred to as TCGA BLCA [@Liu2018; @Robertson2017], 
the IMvigor cohort of chemotherapy-resistant cancers treated subsequently with immunotherapy [@Balar2017], and 
the BCAN cohort of metastatic cancer patients administered systemic chemotherapy [@Damrauer2022].

@@

The following genes coding for FGF, FGFR and FGFBP (FGF-binding proteins) were investigated in the current report [@Korc2009; @Turati2022; @Bennett1995; @Gubbiotti2016; @Mundhenke2002; @Midorikawa2003; @Leali2012; @Xie2020; @Hagihara2000; @Pan2021; @Mohammadi2005; @Hu2009a]:

@@

>>>>

* Five genetic and transcrional collectives of urothelial cancer patients were evaluated: the GENIE BLCA cohort, an urothelial carcinoma subset of the large GENIE project (n = `r nrow(genie$clinic)`) [@Sweeney2017], the predominantly muscle-invasive MSK IMPACT cohort (n = `r nrow(msk$clinic)`) [@Clinton2022], the muscle-invasive TCGA BLCA collective (n = `r nrow(tcga$clinic)`) [@Liu2018; @Robertson2017], the IMvigor cohort of late stage, platinum-refractory cancers subjected to an anti-PD-L1 treatment (n = `r nrow(imvigor$clinic)`) [@Balar2017], and the BCAN cohort of metastatic urothelial cancers (n = `r nrow(bcan$clinic)`) [@Damrauer2022]. Detailed information on somatic mutations and their effects on protein sequence was provided for the GENIE BLCA, MSK IMPACT, TCGA BLCA, and BCAN cohorts. Binarized indexes of presence/absence of somatic mutations in selected genes were available for the IMvigor cohort. Binarized data on gene amplifications and deletions were obtained for the GENIE BLCA, MSK IMPACT, and TCGA BLCA collectives. Gene expression quantification by RNA sequencing and assignment to consensus molecular classes [@Kamoun2020] were retrieved for the TCGA BLCA, IMvigor, and BCAN cohorts. Overall survival information was available for the MSK IMPACT, TCGA BLCA, IMvigor, and BCAN cohorts; data on disease-specific and relapse-free survival was provided for the TCGA BLCA collective. Approximate overall survival was calculated for the GENIE BLCA project participants based on age at enrollment in years and follow-up time in days. Validation of expression of the genes of interest was done with immunohistochemistry data retrieved from the [Human Protein Atlas](https://www.proteinatlas.org/). Characteristic of the investigated collectives is presented in __Table \@ref(tab:tab-cohorts)__.

@@

* At mRNA level, _FGFR3_, _TGFBR3_, and _FGFR1_ were the strongest expressed receptor genes. _FGF7_, _FGF11_, followed by _FGF2_ were the most abundant ligand transcripts; _HSPG2_, _SDC1_, and _CD44_ were the binding protein genes with the highest mRNA counts in the urothelial cancer tissue. Out of `r globals[c("receptors", "ligands", "binding_proteins")] %>% reduce(union) %>% length` genes of interest, `r ncol(hpa$ihc_score) - 1` proteins were quantified by immunohistochemistry (IHC) in the Human Protein Atlas (HPA). The protein quantities were expressed as IHC scores defined as products of staining intensity (1: negative to 4: strong) and quantity of positive cells (1: none to 4: >75%). In the HPA data set, `r `r sum(colMedians(hpa$ihc_score[-1]) > 1)` FGFR, FGF, and FGFBP proteins had median IHC scores $\geq$ 2, and were hence detectable in cancer specimens. At the protein/IHC level, FGFR1, FGFR3, FGFRL1, and FGFR4 were the most abundant receptors. FGF7, FGF10, FGF3, FGF19, and FGF17 were the strongest expressed ligand proteins. SDC2, SDC1, GPC4, CD44, and SDC4 were the most abundant FGFBP proteins (__Figures \@ref(fig:fig-expression)__ - __\@ref(fig:fig-ihc-fgfbp)__, __Table \@ref(tab:tab-expression)__).

@@

flextable::flextable(rep_tabs$cohorts) %>% 
  set_widths(c(2.5, rep(2.3, 6), 2, 2)) %>% 
  fontsize(j = 1:9, size = 9, part = 'all') %>% 
  footnote(1, 8:9, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

## Expression and co-regulation of genes coding for FGFR, FGF, and FGFBP proteins

```{r fig-expression, fig.width = figur::convert(rep_figs$expression, to = 'in')$w, fig.height = figur::convert(rep_figs$expression, to = 'in')$h, fig.cap = 'Expression of genes coding FGFR, FGF, and FGFBP at mRNA and protein level.'}

rep_figs$expression$plot

__Figure \@ref(fig:fig-expression). Expression of genes coding FGFR, FGF, and FGFBP at mRNA and protein level.__

_(A) Transcript levels of FGFR-, FGF-, and FGFBP-coding genes were determined in cancer tissue samples in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` by RNA sequencing and published as $log_2$ gene counts (`r globals$cohort_labs[c("tcga", "imvigor")] %>% collapse_and`) or Z-scores of $log_2$ gene counts (`r globals$cohort_labs["bcan"]`). Median Z-scores of $log_2$ mRNA expression are shown in a heat map. Each tile represents median expression Z-score of a gene in one cohort. Numbers of cancer samples are indicated in the X axis. White tiles represent genes not detected in a given cohort._

_(B) Protein expression of FGFR-, FGF-, and FGFBP-coding genes `r ncol(hpa$intensity) - 1` FGFR-, FGF-, and FGFBP-coding genes in the Human Protein Atlas (HPA) repository was quantified by immunohistochemistry (IHC). The protein expression is shown as IHC scores defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%). Median IHC scores with interquartile ranges are visualized as boxes with whiskers spanning over 150% of the interquartile ranges. Single IHC specimens are visualized as points. Numbers of IHC samples are indicated in the Y axis._

```{r fig-ihc-receptors, fig.width = figur::convert(rep_figs$ihc_receptors, to = 'in')$w, fig.height = figur::convert(rep_figs$ihc_receptors, to = 'in')$h, fig.cap = 'Representative immunohistochemistry staining images for FGFR proteins retrieved from the Human Protein Atlas.'}

list.files('C:/Users/piotr/Desktop/BLCA FGFR/report/figures', 
           pattern = 'ihc_receptor') %>% 
  paste0('C:/Users/piotr/Desktop/BLCA FGFR/report/figures/', .) %>% 
  include_graphics(rel_path = FALSE)

__Figure \@ref(fig:fig-ihc-receptors). Representative immunohistochemistry staining images for FGFR proteins retrieved from the Human Protein Atlas.__

_Images of immunohistochemistry (IHC) staining for FGFR, FGF, and FGFBP in urothelial cancer samples were retrieved from the Human Protein Atlas (HPA). The staining was quantified by the IHC score which was defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%)._
_For each protein, representative images of immunohistochemistry (IHC) staining were selected; the left image corresponds to the minimum IHC score, the middle image shows the staining close to the median IHC score, the right image corresponds to the maximal IHC score._
_The images are labeled with HPA patient identifiers (ID) and IHC scores._

```{r fig-ihc-ligands, fig.width = figur::convert(rep_figs$ihc_ligands, to = 'in')$w, fig.height = figur::convert(rep_figs$ihc_ligands, to = 'in')$h, fig.cap = 'Representative immunohistochemistry staining images for FGF proteins retrieved from the Human Protein Atlas.'}

list.files('C:/Users/piotr/Desktop/BLCA FGFR/report/figures', 
           pattern = 'ihc_ligand') %>% 
  paste0('C:/Users/piotr/Desktop/BLCA FGFR/report/figures/', .) %>% 
  include_graphics(rel_path = FALSE)

__Figure \@ref(fig:fig-ihc-ligands). Representative immunohistochemistry staining images for FGF proteins retrieved from the Human Protein Atlas.__

_Images of immunohistochemistry (IHC) staining for FGFR, FGF, and FGFBP in urothelial cancer samples were retrieved from the Human Protein Atlas (HPA). The staining was quantified by the IHC score which was defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%)._
_For each protein, representative images of immunohistochemistry (IHC) staining were selected; the left image corresponds to the minimum IHC score, the middle image shows the staining close to the median IHC score, the right image corresponds to the maximal IHC score._
_The images are labeled with HPA patient identifiers (ID) and IHC scores._

```{r fig-ihc-fgfbp, fig.width = figur::convert(rep_figs$ihc_fgfbp, to = 'in')$w, fig.height = figur::convert(rep_figs$ihc_fgfbp, to = 'in')$h, fig.cap = 'Representative immunohistochemistry staining images for FGFBP proteins retrieved from the Human Protein Atlas.'}

list.files('C:/Users/piotr/Desktop/BLCA FGFR/report/figures', 
           pattern = 'ihc_fgfbp') %>% 
  paste0('C:/Users/piotr/Desktop/BLCA FGFR/report/figures/', .) %>% 
  include_graphics(rel_path = FALSE)

__Figure \@ref(fig:fig-ihc-fgfbp). Representative immunohistochemistry staining images for FGFBP proteins retrieved from the Human Protein Atlas.__

_Images of immunohistochemistry (IHC) staining for FGFR, FGF, and FGFBP in urothelial cancer samples were retrieved from the Human Protein Atlas (HPA). The staining was quantified by the IHC score which was defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%)._
_For each protein, representative images of immunohistochemistry (IHC) staining were selected; the left image corresponds to the minimum IHC score, the middle image shows the staining close to the median IHC score, the right image corresponds to the maximal IHC score._
_The images are labeled with HPA patient identifiers (ID) and IHC scores._

```{r tab-expression, tab.cap = "Expression of FGFR-, FGF-, and FGFBP-coding genes at mRNA level was quantified in the TCGA BLCA, IMvigor, and BCAN cohort by RNA sequencing. The mRNA amounts were provided as log2-transformed gene counts (TCGA BLCA and IMvigor) or Z-scores of log2 gene counts (BCAN). Protein expression of the genes of interest was measured in the Human Protein Atlas (HPA) by immunohistochemistry (IHC). The protein amounts are metered by the IHC score defined as a product of staining intensity and quantity of positive cells. Median mRNA and protein amounts with interquartile ranges, ranges, and numbers of cancer samples are listed. The table is available as a supplementary Excel file."}

rep_tabs$expression[1, 1] %>% 
  mutate(Variable = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(18)

@@

Five genomic and transcriptomic collectives of urothelial carcinoma patients were evaluated in the current report: 
the urothelial cancer subset of the GENIE cohort referred to as GENIE BLCA [@Sweeney2017], 
the MSK IMPACT collective [@Clinton2022], 
the bladder cancer TCGA collective of muscle-invasive urothelial cancers referred to as TCGA BLCA [@Liu2018; @Robertson2017], 
the IMvigor cohort of chemotherapy-resistant cancers treated subsequently with immunotherapy [@Balar2017], and 
the BCAN cohort of metastatic cancer patients administered systemic chemotherapy [@Damrauer2022].

@@

The following genes coding for FGFR, FGFF, and FGFBP (FGF-binding proteins) were investigated in the current report [@Korc2009; @Turati2022; @Bennett1995; @Gubbiotti2016; @Mundhenke2002; @Midorikawa2003; @Leali2012; @Xie2020; @Hagihara2000; @Pan2021; @Mohammadi2005; @Hu2009a]:

@@

For verification of expression of the FGFR-, FGF-, and FGFBP-coding genes at protein level, we resorted to the [Human Protein Atlas (HPA) repository](https://www.proteinatlas.org/), which stores results of immunohistochemical (IHC) staining. 
Entries of the HPA database were retrieved as XML documents, from which patient information, staining quantification, and URLs to IHC images for the genes of interest was retrieved (R package _xml2_) [@Wickham2023]. 
Protein expression in the IHC samples was expressed as the IHC score defined as a product of staining (1: negative to 4: high), staining intensity (1: negative to 4: strong), and quantity of positive cells (1: none to 4: >75%).
